---
title: "Filters"
output:
  beamer_presentation: default
  ioslides_presentation: default
date: "2024-10-07"
---

```{r setup, include=FALSE}
suppressPackageStartupMessages(library(rjd3toolkit))
library(ggplot2)

knitr::opts_chunk$set(echo = FALSE)
```

## Outline

## Data

### Flow of the river Nule

```{r data, echo = TRUE}
plot(Nile)
```

## Pre-specified filters

Filters based on well-defined mathematical properties

### Henderson filters

```{r henderson, echo = FALSE}

nh<-17

hnile<-rjd3x11plus::henderson(Nile, nh)
ts.plot(ts.union(Nile, hnile), col=c("black", "blue"))

```

## Henderson filter properties


```{r henderson_properties, echo = FALSE}

# helper function
henderson<-function(n){
  jfilter<-rJava::.jcall("jdplus/toolkit/base/core/math/linearfilters/HendersonFilters",     "Ljdplus/toolkit/base/core/math/linearfilters/SymmetricFilter;", "ofLength", as.integer(n))
  w<-rJava::.jcall(jfilter, "[D", "weightsToArray")
  ff<-rjd3filters::moving_average(w, lags=-(length(w)-1)/2)
  return(ff)
}

ff<-henderson(nh)
par(mfrow=c(1,2))
barplot(ff@coefficients, main="weights")
rjd3filters::plot_gain(ff, main="gain")
par(mfrow=c(1,1))
```

## Local polynomials


Regression with a linear trend on 11 periods, using a bi-weight kernel $\alpha(1-x^2)^2)$



```{r lp, echo = FALSE}
lpnile<-rjd3filters::localpolynomials(x=Nile, horizon=5, degree=1, kernel="biweight", endpoints = "DAF")
hnile<-rjd3x11plus::henderson(Nile, nh)
ts.plot(ts.union(Nile, hnile, lpnile), col=c("black", "cyan", "blue"))

```

## Loess filter properties


```{r lp_properties, echo = FALSE}
lp<-rjd3filters::lp_filter(horizon=5, degree=1, kernel="biweight", endpoints = "DAF")
par(mfrow=c(1,2))
barplot(lp@sfilter@coefficients, main="weights")
rjd3filters::plot_gain(lp@sfilter, main="gain")
par(mfrow=c(1,1))

```

## Model-based filters (I)

### Structural model (fixed parameters)

\tiny

\begin{align*} 
y(t) &= T(t)+N(t) \\
\Delta^2 T(t) &= \epsilon_T(t) \quad &\epsilon_T \sim N(0,1) \\
N(t) &=\epsilon_N(t) \quad &\epsilon_N \sim N(0,\lambda)
\end{align*}

```{r sts_code, echo = TRUE}

hp_ucm<-function(lambda=25){
  i2<-rjd3toolkit::arima_model("trend", delta = c(1,-2, 1))
  noise<-rjd3toolkit::arima_model(variance=lambda)
  ucm<-rjd3toolkit::ucarima_model(components=list(i2, noise))
}

ucm_estimate<-function(x, ucm, stdev=TRUE){
  jucm<-rjd3toolkit:::.r2jd_ucarima(ucm)
  jcmps<-rJava::.jcall("jdplus/toolkit/base/r/arima/UcarimaModels", 
            "Ljdplus/toolkit/base/api/math/matrices/Matrix;", "estimate",
            as.numeric(x), jucm, as.logical(stdev))
  return(rjd3toolkit:::.jd2r_matrix(jcmps))
}

hp_estimate<-function(s, lambda=25){
  ucm<-hp_ucm(lambda = lambda)
  rslt<-ucm_estimate(s, ucm)
}

```

\normalsize

## Model-based filters (I)

### Structural model (fixed parameters)


```{r sts_rslt, echo = FALSE}
lambda=25
hp<-hp_estimate(Nile, lambda)
trend=hp[,1]
etrend=hp[,3]

w<-data.frame(date=rjd3toolkit::daysOf(Nile),
              y=as.numeric(Nile),
              trend=trend,
              lp=lpnile,
              h=hnile,
              min=as.numeric(trend-etrend),
              max=as.numeric(trend+etrend))

p<-ggplot(data=w) +
  geom_line(aes(x=date, y=y), color='black') +
  geom_line(aes(x=date, y=trend), color='red') +
  geom_line(aes(x=date, y=lp), color='blue') +
  geom_line(aes(x=date, y=h), color='green') +
  geom_ribbon(aes(x=date, ymin=min, ymax=max), alpha=.2)

plot(p)

```



## Model-based filters properties

### Structural model (fixed parameters)

```{r sts_properties, echo = FALSE}
hp<-hp_ucm(lambda=lambda)
wk_trend<-rjd3toolkit::ucarima_wk(hp, 1)

par(mfrow=c(1,2))
barplot(wk_trend$filter[1:18], main="weights")
nxlab = 7
xlim=c(0, pi)
xlabel <- function(x, symbol = "pi"){
  fracs <- strsplit(attr(MASS::fractions(x), "fracs"), "/")  # convert to fractions
  labels <- sapply(fracs, function(i)
    if (length(i) > 1) { paste(i[1], "*", symbol, "/", i[2]) }
    else { paste(i, "*", symbol) })
  labels <- sub("0 * pi", "0", labels, fixed = TRUE)
  labels <- sub("1 * pi", " pi", labels, fixed = TRUE)
  parse(text = labels)
}

gain<-sqrt(wk_trend$gain2)
n<-length(gain)-1
plot(cbind((0:600)*(pi/n),gain), type = "l",
       xaxt = "n", xlab = "",
       ylab = "gain", xlim = xlim, main="gain")
    x_lab_at <- seq(xlim[1]/pi, xlim[2]/pi, length.out = nxlab)
    axis(1, at = x_lab_at * pi, labels = xlabel(x_lab_at))
par(mfrow=c(1,1))



```
## Model-based filters properties

### Canonical decomposition

\tiny

```{r arima_code, echo = TRUE}

hp_ucm<-function(lambda=25){
  i2<-rjd3toolkit::arima_model("trend", delta = c(1,-2, 1))
  noise<-rjd3toolkit::arima_model(variance=lambda)
  ucm<-rjd3toolkit::ucarima_model(components=list(i2, noise))
}

ucm_estimate<-function(x, ucm, stdev=TRUE){
  jucm<-rjd3toolkit:::.r2jd_ucarima(ucm)
  jcmps<-rJava::.jcall("jdplus/toolkit/base/r/arima/UcarimaModels", 
            "Ljdplus/toolkit/base/api/math/matrices/Matrix;", "estimate",
            as.numeric(x), jucm, as.logical(stdev))
  return(rjd3toolkit:::.jd2r_matrix(jcmps))
}

hp_estimate<-function(s, lambda=25){
  ucm<-hp_ucm(lambda = lambda)
  rslt<-ucm_estimate(s, ucm)
}

```

\normalsize

## Model-based filters

### Canonical decomposition



```{r arima_rslt, echo = FALSE}
arima<-sarima_estimate(Nile, order=c(0,1,1))
sarima<-sarima_model(period = 1, d=1, theta = arima$parameters$val[1])
ucm<-sarima_decompose(sarima)
rslt<-ucm_estimate(Nile, ucm)

trend<-rslt[,1]
etrend<-rslt[,3]
w<-data.frame(date=rjd3toolkit::daysOf(Nile),
              y=as.numeric(Nile),
              trend=trend,
              lp=lpnile,
              h=hnile,
              min=as.numeric(trend-etrend),
              max=as.numeric(trend+etrend))

p<-ggplot(data=w) +
  geom_line(aes(x=date, y=y), color='black') +
  geom_line(aes(x=date, y=trend), color='red') +
  geom_line(aes(x=date, y=lp), color='blue') +
  geom_line(aes(x=date, y=h), color='green') +
  geom_ribbon(aes(x=date, ymin=min, ymax=max), alpha=.2)

plot(p)

```


## Model-based filters properties

### Canonical decomposition


```{r arima_properties, echo = FALSE}
wk_trend<-rjd3toolkit::ucarima_wk(ucm, 1)

par(mfrow=c(1,2))
barplot(wk_trend$filter[1:18], main="weights")

gain<-sqrt(wk_trend$gain2)
n<-length(gain)-1
plot(cbind((0:600)*(pi/n),gain), type = "l",
       xaxt = "n", xlab = "",
       ylab = "gain", xlim = xlim, main="gain")
    x_lab_at <- seq(xlim[1]/pi, xlim[2]/pi, length.out = nxlab)
    axis(1, at = x_lab_at * pi, labels = xlabel(x_lab_at))
par(mfrow=c(1,1))

```
